Cominuscule subvarieties of flag varieties

### 2025-10-24 ms.tex
Add files via upload

\subsection{The Hasse diagram and irreducibles}
\begin{theorem}\label{theorem:G.irreps}
Suppose that \((X,G)\) is a flag variety.
Pick a point \(x_0\in X\), let \(P:=G^{x_0}\).
Let \(G_0\subseteq P\) be a reductive Levi factor containing the Cartan subgroup.
Pick a Cartan subgroup of \(G\) in \(G_0\).
The associated graded Lie algebra \(\agr\LieG\) is the direct sum of irreducible \(P\)-modules
\[
\agr\LieG=\bigoplus_{\bar\aa} \agr[\bar\aa]\LieG.
\]
Each of these \(\agr[\bar\aa]\LieG\) is the sum of the root spaces of a unique component of the Hasse diagram of \(\LieG\).
Conversely every component of the Hasse diagram of \(\LieG\) occurs uniquely in this way.
In particular, the box is the component of the highest root, and its associated \(P\)-module is \(\LieZ\), the center of the unipotent radical \(\LieG_+\) of \(\LieP\).
\end{theorem}
\begin{proof}
Our proof follows Wolf \cite{Landsberg/Manivel:2003} p.~5 Proposition~2.4, \cite{Wolf:2011} p.~300 \S{}8.13.3, who says that it is an unpublished result of Kostant.
Let \(V:=\agr[\bar\aa]\LieG\).
Clearly \(V\) is a \(P\)-module, with trivial \(G_+\)-action, so a \(G_0\)-module, graded by weight:
\[
\gr V=\bigoplus_{\lambda} \gr[\lambda]V.
\]
As \(G_0\)-modules, \(\agr\LieG\) is isomorphic to \(\LieG\), so each \(\gr[\lambda]V\) is a \(G_0\)-submodule of \(\LieG\).
Every root \(\lambda\), as a weight, has a weight space of multiplicity one \cite{Fulton/Harris:1991} p.~198, spanned by a root vector.
Decompose \(V\) into irreducible \(G_0\)-modules
\[
V=\bigoplus_{i\in I}U_i.
\]
Each \(U_i\) has multiplicity one at its weights.
Hence any distinct \(U_i,U_j\) have disjoint weights.
Each \(U_i\) is a \(G_0\)-submodule of \(\LieG\), so its weight vectors are root vectors of \(\LieG\) and \(\LieG_0\) acts on each \(U_i\) by Lie bracket.

>>>>

\subsection{The Hasse diagram and irreducibles}
\begin{theorem}\label{theorem:G.irreps}
Suppose that \((X,G)\) is a flag variety.
Pick a point \(x_0\in X\), let \(P:=G^{x_0}\).
Let \(G_0\subseteq P\) be a reductive Levi factor containing the Cartan subgroup.
Pick a Cartan subgroup of \(G\) in \(G_0\).
The associated graded Lie algebra \(\agr\LieG\) is the direct sum of irreducible \(P\)-modules
\[
\agr\LieG=\bigoplus_{\bar\aa} \agr[\bar\aa]\LieG.
\]
The subgroup \(G_+\subseteq P\) acts trivially on the associated graded, so each is a module of \(G_0\cong P/G_+\).
Each of these \(\agr[\bar\aa]\LieG\) is the sum of the root spaces of a unique component of the Hasse diagram of \(\LieG\).
Conversely every component of the Hasse diagram of \(\LieG\) occurs uniquely in this way.
In particular, the box is the component of the highest root, and its associated \(P\)-module is \(\LieZ\), the center of the unipotent radical \(\LieG_+\) of \(\LieP\).
\end{theorem}
\begin{proof}
Our proof follows Wolf \cite{Landsberg/Manivel:2003} p.~5 Proposition~2.4, \cite{Wolf:2011} p.~300 \S{}8.13.3, who says that it is an unpublished result of Kostant.
Let \(V:=\agr[\bar\aa]\LieG\).
Clearly \(V\) is a \(P\)-module, with trivial \(G_+\)-action, so a \(G_0\)-module, graded by weight:
\[
\gr V=\bigoplus_{\lambda} \gr[\lambda]V.
\]
As \(G_0\)-modules, \(\agr\LieG\) is isomorphic to \(\LieG\), so each \(\gr[\lambda]V\) is a \(G_0\)-submodule of \(\LieG\).
Every root \(\lambda\), as a weight, has a weight space of multiplicity one \cite{Fulton/Harris:1991} p.~198, spanned by a root vector.
Decompose \(V\) into irreducible \(G_0\)-modules
\[
V=\bigoplus_{i\in I}U_i.
\]
Each \(U_i\) has multiplicity one at its weights.
Hence any distinct \(U_i,U_j\) have disjoint weights.
Each \(U_i\) is a \(G_0\)-submodule of \(\LieG\), so its weight vectors are root vectors of \(\LieG\) and \(\LieG_0\) acts on each \(U_i\) by Lie bracket.
